Making ice sheet models \\ scale properly

### 2023-01-21 talk/slides.tex
a bit more talk

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.

\usepackage{empheq,bm,xspace,fancyvrb,soul}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,decorations.markings,decorations.pathreplacing,fadings,positioning}

\usepackage[kw]{pseudo}
\pseudoset{left-margin=15mm,topsep=5mm,idfont=\texttt,st-left=,st-right=}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:
%\beamerdefaultoverlayspecification{<+->}

@@

\begin{itemize}
\item before numerical modeling \dots \emph{what} are we modeling?
\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
\end{itemize}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glaciers lie on \emph{topography}, and sometimes they float  (\emph{ice shelf})
\item glacier geometry and velocity evolve \emph{in contact with climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt when floating
    \item[$\circ$] calving into the ocean
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming model performance analysis, I will ignore quite a bit of physics
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \item[$\circ$] solid earth deformation
    \end{itemize}

@@

\smallskip
\item<2-> {\small map: \quad $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item the \alert{data} $a(t,x,y)$, $b(x,y)$  is defined on a fixed domain:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2> the solution \alert{surface elevation} $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] $s=b$ where there is no ice
    \end{itemize}

\medskip
\item<2> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$ on which the solution \alert{velocity} $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

\vspace{-2mm}
\end{itemize}
\end{column}
\begin{column}{0.4\textwidth}
\includegraphics<1>[width=\textwidth]{images/domain-data.png}

\includegraphics<2>[width=\textwidth]{images/domain-velocity.png}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{mass and momentum conservation}

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
    \end{itemize}
\item a key idea is that conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

    \item[$\circ$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
    \end{itemize}

@@

\begin{itemize}
\item ice sheet evolution is a \alert{free-boundary} problem for conserved mass and momentum quantities
\item specifically, the surface kinematic equation
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the (fixed, map-plane) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

@@

\begin{frame}{basic ice sheet model: strong form}

\begin{itemize}
\item nonlinear complementarity problem (NCP) \onslide<2->{coupled to Stokes}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}}
\end{align*}

    \begin{itemize}
    \item $\bu|_s=0$ where no ice
    \item<2-> viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{basic ice sheet model: its a DAE system anyway}

\begin{itemize}
\item for clarity and simplicity, declutter by forgetting complementarity and all the other boundary conditions, too \dots
\item get surface kinematic equation coupled to Stokes:
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 && \text{on } \Omega \subset \RR^2 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''}
\end{align*}
\item observe that $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
\item \emph{however}, only the first of the 5 equations has a time derivative
\item the problem is a \alert{differential algebraic equation} system, e.g.
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item[$\circ$] \dots actually in infinite dimensions (\emph{PDAE}?), and subject to complementarity (\emph{free boundary problem for PDAE})
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{important context (current research)}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this ``NCP coupled to Stokes'' glacier evolution model
    \begin{itemize}
    \item[$\circ$] however, I think most researchers would agree this is the intended model!
    \item[$\circ$] well-posedness of shallow approximations has been considered\footnote{e.g.~Jouvet \& Bueler (2012)}
    \end{itemize}

\medskip
\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution (\emph{time-split} or \emph{explicit time-stepping}),
\item<2> and glacier evolution is addressed with minimal awareness of complementarity
\end{itemize}
\end{frame}

\begin{frame}{the mass-continuity equation view}

\begin{itemize}
\item thickness-transport form helps expose evolution character
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

\smallskip
\item<3-> \emph{question:} is it really an advection?
\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\end{itemize}
\end{frame}

\begin{align*}
\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item both forms are highly-nonlinear: \, $\bU=\bU(H,\grad s), \, D = D(H,\grad s)$
\item the diffusion form is only \emph{literally} valid in the lubrication approximation (more below)
\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences} regardless
\end{itemize}
\end{frame}

\begin{frame}{numerical shallow ice sheet models}

\begin{itemize}
\item the ``basic ice sheet model'' (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
    \begin{itemize}
    \item[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
    \item[$\circ$] goal: \alert{make it fast enough to use!}
    \end{itemize}

\smallskip
\item<2-> current-technology models use various shallow simplifications
    \begin{itemize}
    \item<2->[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
    \end{itemize}
\item<2-> the simplest shallow approximation is a lubrication approximation, the \alert{shallow ice approximation} (SIA)
\item<2-> in the SIA, $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator

\begin{frame}{shallow ice approximation NCP: details}

{\small
\begin{align*}
s - b &\ge 0 \\
\frac{\partial s}{\partial t} + \Phi(s) - a &\ge 0 \\
(s - b) \left(\frac{\partial s}{\partial t} + \Phi(s) - a\right) &= 0
\end{align*}
\begin{equation*}
\Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
\end{equation*}
}

\begin{itemize}
\item constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation

\medskip
\item<2> thus: $\Phi(s) = - \bu|_s \cdot \bn_s$ resolves to a \alert{doubly-nonlinear} differential operator, of porous medium and $\pp$-Laplacian type simultaneously
\item<2> Jouvet \& Bueler (2012): existence for this problem
\end{itemize}
\end{frame}

@@

@@

{\footnotesize
\begin{align*}
\text{advective ice sheet model schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion ice sheet model schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
}

\bigskip
\begin{itemize}
\item recall: traditional numerical analysis applied to these equations
\end{itemize}

@@

    \begin{itemize}
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \item[$\circ$] maximum time step from CFL using $\max |\bU|$
    \end{itemize}
}
\only<3>{
\item we expect \alert{implicit} time stepping for diffusions
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (H_{j+1}^{\ell+1} + H_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (H_{j}^{\ell+1} + H_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] or use: Crank-Nicolson, BDF, \dots
    \end{itemize}
}
\end{itemize}
\end{minipage}
\end{frame}

@@

\begin{itemize}
\item current-technology models mostly ``believe'' the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] an embarrassing situation \dots the problem is a DAE!
    \item[$\circ$] performance consequences: stability is determined by the coupled solution velocity $\bU$ not a pre-determined velocity
    \end{itemize}

\medskip
\item<2> an alternative should be obvious to any numerical analyst \dots
\item<2> \alert{implicit time-stepping} is expected for DAE problems
\item<2> implicit steps imply a sequence of problems (``equations'') to solve at each time step
\item<2> here: implicit $\implies$ NCP coupled to Stokes at each time step
\end{itemize}
\end{frame}

@@

@@

FIXME explain in terms of 2D Poisson
\end{frame}

@@

FIXME Jouvet et al.~2021
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
\item<3-> \alert{coming soon} from current research:
   \begin{itemize}
   \item[$\circ$] scalable Stokes solvers (Isaac et al.~2015)
   \item[$\circ$] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[$\circ$] ML emulators (Jouvet et al.~2021)
   \end{itemize}
\item<4-> scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}

>>>>

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{empheq,bm,xspace,fancyvrb,soul}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,decorations.markings,decorations.pathreplacing,fadings,positioning}
\usepackage[kw]{pseudo}
\pseudoset{left-margin=15mm,topsep=5mm,idfont=\texttt,st-left=,st-right=}

@@

\begin{itemize}
\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
\end{itemize}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glaciers lie on \emph{topography}, and sometimes they float  (\emph{ice shelf})
\item glacier geometry and velocity evolve \emph{in contact with climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] sub-shelf melt (when floating)
    \item[$\circ$] calving (into ocean)
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming model performance analysis, I will ignore much of glacier physics
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \item[$\circ$] solid earth deformation
    \end{itemize}

@@

\smallskip
\item {\small \alert{map: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$} }
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$  is defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice!
    \end{itemize}
\item<3> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which (solution) velocity $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

\vspace{-2mm}
\end{itemize}
\end{column}
\begin{column}{0.46\textwidth}
\includegraphics<1>[width=\textwidth]{images/domain-data.png}

\includegraphics<2>[width=\textwidth]{images/domain-surface.png}

\includegraphics<3>[width=\textwidth]{images/domain-velocity.png}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{basic ice sheet model: conservation}

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
    \end{itemize}

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

        \begin{itemize}
        \item[$\vartriangleright$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
        \end{itemize}
    \end{itemize}

@@

\begin{itemize}
\item ice sheet evolution is a \alert{free-boundary} problem for conserved quantities
\item specifically, the surface kinematic equation
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the (fixed) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

@@

\begin{frame}{basic ice sheet model: strong form = NCP coupled to Stokes}

\begin{itemize}
\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to Stokes}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t)$}}
\end{align*}

    \begin{itemize}
    \item[$\circ$] $\bu|_s=0$ where no ice
    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{basic ice sheet model: is a DAE system}

\begin{itemize}
\item for this slide, forget complementarity and all boundary conditions
\item result is surface kinematic equation coupled to Stokes:
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

\vspace{-2mm}
    \begin{itemize}
    \item[$\circ$] $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
    \end{itemize}
\item<2-> only the first of these 5 equations has a time derivative
\item<3> the problem is a \alert{differential algebraic equation} (DAE) system:
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item<3>[$\circ$] in infinite dimensions (PDAE?), and subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{basic ice sheet model: current research and numerics}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this basic ice sheet model
    \begin{itemize}
    \item[$\circ$] but most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
    \item[$\circ$] well-posedness of shallow approximations \emph{has} been considered
    \end{itemize}

\medskip
\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution
    \begin{itemize}
    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is taken for granted
    \end{itemize}
\item<2> glacier evolution is addressed with minimal awareness of complementarity
\end{itemize}
\end{frame}

@@

\begin{frame}{ice sheet models: the mass-continuity equation view}

\begin{itemize}
\item thickness transport form helps expose evolution character
\item define
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

\smallskip
\item<3-> \emph{question:} is it really an advection?
\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H,$$
and ice is slow (DAE) so there are no characteristics
\end{itemize}
\end{frame}

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\item the diffusion schema is only literal in the lubrication approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}, regardless of your advective schema preference
\end{itemize}
\end{frame}

\begin{frame}{shallow ice sheet models}

\begin{itemize}
\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
    \begin{itemize}
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}

\smallskip
\item current-technology models use various shallow simplifications
    \begin{itemize}
    \item[$\circ$] usual justification: ``Stokes is too expensive at scale''
    \end{itemize}
\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)

\begin{frame}{shallow ice approximation NCP}

\begin{itemize}
\item in the shallow ice approximation, the surface motion contribution $\Phi(s) = - \bu|_s \cdot \bn_s$ has a formula:
{\small
\begin{gather*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0 \\
{\color{FireBrick} \Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)}
\end{gather*}

\vspace{-2mm}
    \begin{itemize}
    \item[$\circ$] constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
    \end{itemize}
}

\medskip
\item<2> $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item<2>[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item<2>[$\circ$] \emph{local} in surface and bed topography
    \item<2>[$\circ$] existence is known for this problem
         \begin{itemize}
         \item[$\vartriangleright$] Jouvet \& Bueler (2012); NCP is written as a \alert{variational inequality} weak form
         \end{itemize}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{nonlocality}

\begin{itemize}
\item the basic ice sheet model (NCP coupled to Stokes) problem actually has a \emph{non}-local surface velocity function $\Phi(s) = - \bu|_s \cdot \bn_s$
{\small
\begin{equation*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
\end{equation*}
}
\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the shallow ice approximation velocity responds only underneath the surface perturbation}
\end{itemize}

\begin{center}
\includegraphics<1>[width=0.6\textwidth]{images/stokes-greens-arndt.png}
\includegraphics<2>[width=0.6\textwidth]{images/sia-greens-arndt.png}
\end{center}
\end{frame}

@@

{\footnotesize
\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
}

\bigskip
\begin{itemize}
\item recall traditional numerical analysis applied to these equations
\end{itemize}

@@

    \begin{itemize}
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \item[$\circ$] conditional stability
    \item[$\circ$] maximum time step from CFL using $\max |\bU|$
    \end{itemize}
}
\only<3>{
\item we expect \alert{implicit} time stepping for diffusions
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] or: Crank-Nicolson, BDF, \dots
    \item[$\circ$] unconditional stability
    \item[$\circ$] must solve equations at each step
    \end{itemize}
}
\end{itemize}
\end{minipage}
\end{frame}

@@

\begin{itemize}
\item current researchers mostly believe the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] embarrassing: the problem is a DAE
    \item[$\circ$] performance and design consequences: stability is determined by the coupled solution velocity $\bU$
    \end{itemize}

\medskip
\item<2> \alert{implicit time-stepping} is expected for DAE problems
    \begin{itemize}
    \item[$\circ$] the obvious alternative
    \end{itemize}
\item<2> a sequence of problems (``equations'') must be solved at each time step: NCP coupled to Stokes
\end{itemize}
\end{frame}

@@

\begin{frame}{the Stokes problem for ice}

\begin{itemize}
\item from now on assume our ice sheet model is \emph{not} shallow
\item whether time steps are explicit or implicit, a Stokes problem is solved at each step
\item for a given icy domain $\Lambda \subset \RR^3$:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item how fast is the numerical solution process?
\end{itemize}
\end{frame}

@@

FIXME explain for Poisson
\end{frame}

@@

FIXME Jouvet et al.~2021

\begin{center}
\includegraphics[width=0.45\textwidth]{images/cnn-jouvet.png}

\includegraphics[width=0.65\textwidth]{images/tiles-jouvet.png}
\end{center}
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress will be made in all of these areas
   \end{itemize}
\item<3-> \alert{coming soon} from current research:
   \begin{enumerate}
   \item[1.] machine learning emulators (Jouvet et al.~2021)
   \item[2.] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
   \end{enumerate}
\item<4-> scalable solvers for the implicit-step problem require advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}
